---
title: "Analysis of bike sharing data"
author: "Elise Gourri, Emilia Marlene Ribeiro Peixoto, Pascal Albisser"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: TRUE
    code_folding: hide
---

TEST COMMIT because the last one was not complete...

```{r libraries, message = FALSE, echo = FALSE}
library(kableExtra)
library(ggplot2)
library(dplyr)
library(data.table)
library(tidyr) 
library(scales)
library(tidyverse)
library(kernlab)
library(e1071)
library(caret)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = function() {
  if (knitr::is_latex_output())
    "latex" else "pipe"
})
```

# Dataset

Our Dataset, pls run `scripts/download-data.R` first.

```{r head_data}
data <- fread("data/trips_s.csv")
kable(head(data))
```


```{r trips_per_city}
data %>% 
  ggplot(mapping = aes(x = city, fill = city)) + 
    geom_bar() +
  geom_text(aes(label = after_stat(count)), stat = 'count', nudge_y = 0.5,
    colour = 'black',size = 3) +
    labs(title = "Number of trips per city")
```


```{r trips_per_month, message = FALSE}
data %>% 
  group_by(city, year_month = paste(format(started_at, "%Y-%m"), "-01", sep="")) %>% 
  summarise(n = n()) %>% 
  ggplot(mapping = aes(y = n, x = as.Date(year_month), group = city, colour = city)) + 
    geom_line() +
    geom_point() +
    labs(title = "Sum of trips per city and month", y = "number of trips", x = "")
```

## Duration of trips

__Emilia told me (Elise) that some trip durations are absurd. We need to check that!__

# Weather (Elise)

I started using our main dataset but because I want to summarize data per day. 
I don't know how to group the data because we have a different amount of data per hour 
with the bike rental data...
So for now I will try with the weather_data dataframe. We can ditch all this later if 
we don't use it.

## Precipitations

Here, I am checking if the na in our weather data precipitations are also na on the metesostat.net website.
I checked 2021-08-30 3h Oslo: on the website, it is everywhere 0. So we should be fine replacing the NA by 0 for precipitation.

```{r precip-na}
precip_na <- filter(weather_data,is.na(weather_data$prcp)) %>% arrange(desc(date))
head(precip_na,500)
```

```{r precipitations, message = FALSE}
precip <- weather_data %>% filter(date > "2017-12-31") %>%
  mutate(prcp = replace_na(prcp, 0)) %>%
  group_by(city, year_month = as.Date(paste(format(date, "%Y-%m"), "-01", sep=""))) %>% 
  summarise(precipitations = sum(prcp)) 

ggplot(data = precip, mapping = aes(y = precipitations, x = year_month, group = city, colour = city)) + 
  geom_point() +
  geom_line() +
  labs(title = "Sum of precipitations per city and month", y = "precipitations in mm", x = "")
```


## Temperature

Here, I am checking if the na in our weather data temperature are also na on the metesostat.net website.
```{r temperature-na}
temp_na <- filter(weather_data,is.na(weather_data$temp)) %>% arrange(desc(date))
length(temp_na$date)
temp_na %>% group_by(city,date) %>% count() %>% arrange(desc(n))
head(temp_na,200)
```

There are 1'404 missing values which we could fill with fill(weather_data, temp, .direction = 'down').
I don't know how many bike data are concerned. Another possibility, would be to just remove the bike data with missing temperature.
If we choose to fill missing data, we would need to do it before merging the weather and bike datasets.
Let me know what you guys think.

```{r temperature}
tempe <- weather_data %>% filter(date > "2017-12-31") %>%
  #mutate(temp = replace_na(temp, 0)) %>%
  fill(temp, .direction = 'down') %>%
  group_by(city, year_month = as.Date(paste(format(date, "%Y-%m"), "-01", sep=""))) %>% 
  summarise(temp_av = mean(temp)) 

ggplot(data = tempe, mapping = aes(y = temp_av, x = year_month, group = city, colour = city)) + 
  geom_point() +
  geom_line() +
  labs(title = "Average temperature per city and month", y = "temperature in °C", x = "")
```

# Linear Model

# Generalised Linear Model with family set to Poisson

# Generalised Linear Model with family set to Binomial

# Generalised Additive Model

# Neural Network

# Support Vector Machine

We predict the city based on latitude and longitude.

```{r cities_dataset}
# Cities dataset
cities_start <- data %>% select(c(start_station_latitude, start_station_longitude, city))
colnames(cities_start) <- c("latitude", "longitude","city")

cities_end <- data %>% select(c(end_station_latitude, end_station_longitude, city))
colnames(cities_end) <- c("latitude", "longitude","city")

cities <- rbind(cities_start, cities_end)
rm(cities_start, cities_end)

cities$city <- as.factor(cities$city)

colSums(is.na(cities)) # no missing values!

# form training and testing datasets
set.seed(10)
indices <- createDataPartition(cities$city, p=0.7, list=F)

train_cities <- cities %>% slice(indices)
test_cities <-  cities %>% slice(-indices)
test_cities_truth <- cities %>% slice(-indices) %>% pull(city)
```

## Linear kernel

```{r SVM_city_linear}
cities_svm <- svm(city ~. , train_cities, kernel = "linear", scale = TRUE, cost = 10)

#make predictions
test_pred <- predict(cities_svm, test_cities)
table(test_pred)
#evaluate the results
conf_matrix <- confusionMatrix(test_pred, test_cities_truth)
conf_matrix

#plot does not work...
#plot(cities_svm, train_cities, start_station_latitude, start_station_longitude)
```

Only 3 data points were misclassified. We built a SVM with a prediction accuracy of nearly 100%.
__cf. Iris lab: We should do some cross-validation to ensure that we are not simply lucky with the train/test split. But on the other end... it is good enough, no?__

## Radial kernel

```{r SVM_city_radial}
cities_svm_radial <- svm(city ~. , train_cities, kernel = "radial", scale = TRUE, cost = 100)

#make predictions
test_pred2 <- predict(cities_svm_radial, test_cities)
table(test_pred2)
#evaluate the results
conf_matrix2 <- confusionMatrix(test_pred2, test_cities_truth)
conf_matrix2
```
Same results with another type of kernel.

__It would be useful to identify those 3 wrong values, but I don't know how...__

# Solve an optimisation problem
